Gender disparity in computational biology research publications

### 2017-02-14 manuscript.tex
fixing formatting so tex doc can build

\documentclass[11pt]{article}
\usepackage{fullpage}
\title{Gender disparity in computational biology research publications}
\author[1][2]{Kevin S. Bonham, PhD\thanks{kevin_bonham@hms.harvard.edu}}
\author[3]{Melanie I. Stefan, PhD}
\affil[1]Microbiology and Immunobiology, Harvard Medical School. Boston, MA, USA
\affil[2]Curriculum Fellows Program and Educational Laboratory, Harvard Medical School. Boston, MA, USA
\affil[3]Centre for Integrative Physiology, Edinburgh Medical School: Biomedical Sciences, University of Edinburgh. Edinburgh, United Kingdom

\begin{document}

\section{Abstract}
While women are generally underrepresented in STEM fields, there are noticeable differences between fields. For instance, the gender ratio in biology is more balanced than in computer science. We were interested in how this difference is reflected in the interdisciplinary field of computational/quantitative biology. To this end, we examined the proportion of female authors in publications from the PubMed and arXiv databases. There are fewer female authors on research papers in computational biology, as compared to biology in general. This is true across authorship position, year, and journal impact factor. A comparison with arXiv shows that quantitative biology papers have a higher ratio of female authors than computer science papers, placing computational biology in between its two parent fields in terms of gender representation. Both in biology and in computational biology, a female last author increases the probability of other authors on the paper being female, pointing to a potential role of female PIs in influencing the gender balance.

@@

We began by analyzing the gender representation in primary publications from 1997 to 2014. Consistent with previous publications, women were substantially less likely to be in senior author positions than first author positions in publications labeled with the Biology (Bio) MeSH term (Last author, $P_{female}$ = 0.245, 95\% CI = [0.243, 0.247], First author, $P_{female}$ = 0.376, 95\% CI = [0.373, 0.378]). (Figure 1A, Table 1). We observed the same trend in papers labeled with the computational biology (comp) MeSH term, though the $P_{female}$ at every author position was 4-6  percentage points lower. An analysis of publications by year suggests that the gender gaps in both biology and computational biology are narrowing, but by less than 1 percentage point  per year (for bio, Δ$P_{female}$ = 0.0035 / year, 95\% CI = [0.0030, 0.0040]. For comp, Δ$P_{female}$ = 0.0049 / year, 95\%CI = [0.0041, 0.0057]). However, the discrepancy between biology and computational biology has been consistent over time (Figure 1B).

Table 1
\begin{tabular}{l l c c c}
Table 1 & & & \multicolumn{2}{c}{95\% CI} \\
Dataset & Position & Mean & Lower & Upper
\multirow{5}{c}{bio} & first & 0.376 & 0.373 & 0.378 \\
& second & 0.379 & 0.376 & 0.381 \\
& other & 0.368 & 0.367 & 0.370 \\
& penultimate & 0.279 & 0.277 & 0.282 \\
& last & 0.245 & 0.243 & 0.247 \\
\multirow{5}{c}{comp} & first & 0.316 & 0.312 & 0.320 \\
& second & 0.322 & 0.317 & 0.327 \\
& other & 0.331 & 0.328 & 0.333 \\
& penultimate & 0.236 & 0.231 & 0.241 \\
& last & 0.207 & 0.203 & 0.211
\end{tabular}

@@

Table 2
Male Last Author
Female Last Author
Dataset
Position
Mean
95\% CI
Mean
95\% CI
Lower
Upper
Lower
Upper
bio
first
0.362
0.359
0.365
0.478
0.472
0.484
second
0.359
0.357
0.362
0.46
0.454
0.466
other
0.355
0.353
0.357
0.425
0.421
0.428
penultimate
0.259
0.256
0.263
0.336
0.330
0.343
comp
first
0.305
0.300
0.311
0.390
0.378
0.402
second
0.306
0.300
0.312
0.379
0.366
0.392
other
0.321
0.318
0.324
0.368
0.361
0.376
penultimate
0.223
0.218
0.229
0.263
0.249
0.277

@@

Women were more likely to be authors in quantitative biology than in computer science in first, second, and middle author positions (Figure 3A, Table 3), though the conventions for determining author order are not necessarily the same in computer science. Nevertheless, women had higher representation in quantitative biology than in computer science for all years except 2009 (Figure 3B). Interestingly, there is a slight but significant (p < 0.05) increase in the proportion of female authors over time in quantitative biology, while there’s no significant increase in computer science preprints.
Table 3

95\% CI
Dataset
Position
Mean
Lower
Upper
arxivbio
first
0.184
0.178
0.190
second
0.210
0.200
0.219
other
0.265
0.253
0.276
penultimate
0.196
0.183
0.209
last
0.148
0.141
0.155
arxivcs
first
0.157
0.155
0.160
second
0.175
0.172
0.179
other
0.188
0.182
0.195
penultimate
0.175
0.170
0.181
last
0.155
0.153
0.158

@@

>>>>

\documentclass[11pt]{article}
\usepackage{fullpage}
\usepackage{multirow}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{authblk}

\title{Gender disparity in computational biology research publications}
\author[1]{Kevin S. Bonham, PhD}
\author[2]{Melanie I. Stefan, PhD}
\affil[1]{Microbiology and Immunobiology, Harvard Medical School. Boston, MA, USA}
\affil[2]{Centre for Integrative Physiology, Edinburgh Medical School. Biomedical Sciences, University of Edinburgh. Edinburgh, United Kingdom}

\renewcommand\Authands{ and }

\begin{document}
\maketitle
\section{Abstract}
While women are generally underrepresented in STEM fields, there are noticeable differences between fields. For instance, the gender ratio in biology is more balanced than in computer science. We were interested in how this difference is reflected in the interdisciplinary field of computational/quantitative biology. To this end, we examined the proportion of female authors in publications from the PubMed and arXiv databases. There are fewer female authors on research papers in computational biology, as compared to biology in general. This is true across authorship position, year, and journal impact factor. A comparison with arXiv shows that quantitative biology papers have a higher ratio of female authors than computer science papers, placing computational biology in between its two parent fields in terms of gender representation. Both in biology and in computational biology, a female last author increases the probability of other authors on the paper being female, pointing to a potential role of female PIs in influencing the gender balance.

@@

We began by analyzing the gender representation in primary publications from 1997 to 2014. Consistent with previous publications, women were substantially less likely to be in senior author positions than first author positions in publications labeled with the Biology (Bio) MeSH term (Last author, $P_{female}$ = 0.245, 95\% CI = [0.243, 0.247], First author, $P_{female}$ = 0.376, 95\% CI = [0.373, 0.378]). (Figure 1A, Table 1). We observed the same trend in papers labeled with the computational biology (comp) MeSH term, though the $P_{female}$ at every author position was 4-6  percentage points lower. An analysis of publications by year suggests that the gender gaps in both biology and computational biology are narrowing, but by less than 1 percentage point  per year (for bio, change in $P_{female}$ = 0.0035 / year, 95\% CI = [0.0030, 0.0040]. For comp, change in $P_{female}$ = 0.0049 / year, 95\%CI = [0.0041, 0.0057]). However, the discrepancy between biology and computational biology has been consistent over time (Figure 1B).

Table 1

@@

Table 2

@@

Women were more likely to be authors in quantitative biology than in computer science in first, second, and middle author positions (Figure 3A, Table 3), though the conventions for determining author order are not necessarily the same in computer science. Nevertheless, women had higher representation in quantitative biology than in computer science for all years except 2009 (Figure 3B). Interestingly, there is a slight but significant (p < 0.05) increase in the proportion of female authors over time in quantitative biology, while there’s no significant increase in computer science preprints.

Table 3

@@

\end{document}
